Copulas and their potential for ecology

### 2019-09-30 Paper.Rmd
just about done editing Q1 methods and results for this round, just want to go through it one more time after Shyamolina makes some changes I requested in an email

<!--***DAN: to do for dealing with multivariate results
1) Here is all you want to present for each system
a) number of total pairs and nonindep pairs
b) number of pairs for which non-N was best; skip AICw values, i.e., rows 4 and 5 of Table 5
c) number of well-fitting locs
d) also keep rows 7-9 of table 5

Get rid of the statement in the methods that says we should not try to combine
model selection results with p-values

Delete the following stuff from sup mat:
DONE Figures S24-27
DONE Tables S7-14, keep Table S15
DONE Tables S16-23, keep S24
DONE S25-32, keep S33
DONE S34-42, keep S42

DONE You will have to go back to the methods and revise those parts
-->

We present green spruce aphid abundance (Data section; Table \ref{tab_data_info}) results
as an example multivariate analysis.
Independence was rejected for each pair of the 10 sampling locations. 
Best-fitting (lowest-AIC) copulas were non-normal
for the large majority of location pairs (72 of 90 pairs), 
<!--***DAN: Shyamolina, please make these numbers autolinks-->
and the AIC
of the normal copula minus the minimum AIC, averaged across 
location pairs, was $2.732$.
<!--***DAN: Shyamolina, please make this an autolink, so the value is
automatically filled in even if it changes-->
Whereas $2.732$
<!--***DAN: Shyamolna, again, autolink, please-->
would be only a marginally meaningful AIC difference for a single
location pair, it is more meaningful as an average across many
pairs. Relatedly, and illustrating the concept here, 
the chances of getting 72 or more 
non-normal location pairs if the chances were equal of getting
a normal versus a non-normal result (and taking into account that
the location pair (A,B) will necessarily 
produce the same result as the pair
(B,A)) is low, $`r 1-pbinom(72/2-1,45,.5)`$. 
<!--***DAN: Shyamolina, please replace the 72 in the above expression
by an autolink, and make this show
3 significant digits, i.e., it should show 3.29e-5, not 3.287332e-05-->
These results help answer the first part of Q1: non-normal
copula structure appears to be meaningfully common in
these data, even if not universal.
Goodness of fit tests in every case 
failed to reject the hypothesis that the AIC-best copula 
family was also 
an objectively adequate description of 
the data; i.e., the collection of 
copula families we used was sufficiently broad to characterize 
these data. Model-averaged lower- and upper-tail
dependence statistics had $2.5$th and $97.5$th quantiles
$(0.123,0.721)$ and $(0.002,0.41)$, respectively,
<!--***DAN: Shyamolina, autolink pls -->
thereby commonly differing 
from what a normal copula would give (i.e., 0), and helping to
answer the second
part of Q1: these data have greater tail dependence 
(lower and upper) than would be expected from a normal-copula null hypothesis. We note however that our model selection 
procedures again do not reveal whether tail-dependence 
parameters are significantly different from $0$, and we refer 
the reader to nonparametric results below for information
on significance. 
Model-averaged lower- minus upper-tail dependence statistics 
were positive for all but a few location pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`). The chances
of getting `r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
or more positive values, here, under a null hypothesis of equal chances 
for positive and negative values (and again accounting for the fact that location pairs (A,B) and (B,A) will show the same result) was
again low, $`r 1-pbinom(86/2-1,45,.5)`$. 
<!--***DAN: Shyamolina, please autolink this, and use 3 sig figs as 
before-->
Thus the spatial synchrony of rarity in the green spruce aphid is stronger 
than the spatial synchrony of outbreaks. This also answers the third part of Q1.

<!--***DAN: got to here. Next step, put in the actual nonparam stats results which are currently in the Appendices-->

Nonparametric statistics verified that tail dependence was asymmetric for the green 
spruce aphid abundance data (Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}).

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes Q1 results for
multivariate data. Results supported the 
conclusions that non-normal copula structure, non-normal tail dependence, and asymmetric tail 
dependence were common, answering Q1 in the affirmative for these data. 
Most location pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copulas that were not the normal copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
better than those of normal copulas (Table \ref{tab_multivar_summary}, rows 4-5).
Best-fitting copulas were nearly always considered an adequate fit 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane-flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
Our 
affirmative, empirically based answer to Q1 for the multivariate datasets is 
represented in Fig. \ref{MasterFigure} as solid outlines around the names
of those datasets, in the middle row of boxes.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than
\textit{Ceratium furca} (results not shown); results supported the conclusion that 
non-normal copula structure and tail dependence, and asymmetry of tail dependence are 
common.

>>>>

We present green spruce aphid abundance (Data section; Table \ref{tab_data_info}) results
as an example multivariate analysis.
Independence was rejected for each pair of the 10 sampling locations. 
Best-fitting (lowest-AIC) copulas were non-normal
for the large majority of location pairs (72 of 90 pairs), 
<!--***DAN: Shyamolina, please make these numbers autolinks-->
and the AIC
of the normal copula minus the minimum AIC, averaged across 
location pairs, was $2.732$.
<!--***DAN: Shyamolina, please make this an autolink, so the value is
automatically filled in even if it changes-->
Whereas $2.732$
<!--***DAN: Shyamolna, again, autolink, please-->
would be only a marginally meaningful AIC difference for a single
location pair, it is more meaningful as an average across many
pairs. Relatedly, and illustrating the concept here, 
the chances of getting 72 or more 
non-normal location pairs if the chances were equal of getting
a normal versus a non-normal result (and taking into account that
the location pair (A,B) will necessarily 
produce the same result as the pair
(B,A)) is low, $`r 1-pbinom(72/2-1,45,.5)`$. 
<!--***DAN: Shyamolina, please replace the 72 in the above expression
by an autolink, and make this show
3 significant digits, i.e., it should show 3.29e-5, not 3.287332e-05-->
These results help answer the first part of Q1: non-normal
copula structure appears to be meaningfully common in
these data, even if not universal.
Goodness of fit tests in every case 
failed to reject the hypothesis that the AIC-best copula 
family was also 
an objectively adequate description of 
the data; i.e., the collection of 
copula families we used was sufficiently broad to characterize 
these data. Model-averaged lower- and upper-tail
dependence statistics had $2.5$th and $97.5$th quantiles
$(0.123,0.721)$ and $(0.002,0.41)$, respectively,
<!--***DAN: Shyamolina, autolinks pls -->
thereby commonly differing 
from what a normal copula would give (i.e., 0), and helping to
answer the second
part of Q1: these data have greater tail dependence 
(lower and upper) than would be expected from a normal-copula null hypothesis. We note however that our model selection 
procedures again do not reveal whether tail-dependence 
parameters are *significantly* different from $0$, and we refer 
the reader to nonparametric results below for information
on significance. 
Model-averaged lower- minus upper-tail dependence statistics 
were positive for all but a few location pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`). The chances
of getting `r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
or more positive values, here, under a null hypothesis of equal chances 
for positive and negative values (and again accounting for the fact that location pairs (A,B) and (B,A) will show the same result) was
again low, $`r 1-pbinom(86/2-1,45,.5)`$. 
<!--***DAN: Shyamolina, please autolink this, and use 3 sig figs as 
before-->
Thus the spatial synchrony of rarity in the green spruce aphid is stronger 
than the spatial synchrony of outbreaks. This answers the third part of Q1.

Nonparametric statistics verified that tail associations were asymmetric for the green 
spruce aphid abundance data. Mean values across all pairs of 
locations of the Spearman and Kendall correlations and the statistics 
$\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, $\Dsq_l$, and $\Dsq_u$ were positive and
confidence intervals excluded $0$ (Table \ref{tab_count_resamp}). 
Mean values of the statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$
also were always positive and had confidence intervals that 
excluded $0$ 
(Table \ref{tab_count_resamp}).

<!--***DAN: Shyamolina, can you please insert code to make what
is currently Table S7 appear here? Please also make it disappear from
the sup mat. I want to move it from there to here, please. I am
talking about the table with label tab_count_resamp. Please also
search both the main text and the supp mat for that label, and 
fix references - you may have to fix some prefixes (SM-, MT-), since
the table will have moved from sup mat to main text.-->

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes Q1 results for
multivariate data. Results supported the 
conclusions that non-normal copula structure, non-normal tail dependence, and asymmetric tail 
dependence were common, though not universal, answering 
Q1 in the affirmative for these data. 
Most location pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copulas that were not the normal copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC values of best-fitting copulas were, on average
across location pairs, between $2.714$ (*Ceratium furca* abundance data)
and $5.764$ (leaf-curling plum aphid first-flight data) lower than
AIC values for the normal copula (Table \ref{tab_multivar_summary}, 
row 4). Best-fitting copulas were nearly always considered an adequate fit 
(Table \ref{tab_multivar_summary}, row 5). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane-flux) showed stronger 
lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 6-8), whereas leaf-curling plum aphid first flight
data showed the reverse. All of the values in Table \ref{tab_multivar_summary}, rows 8 deviated highly significantly
from what would have been expected under a null hypothesis 
of equal chances for positive and negative values.

Asymmetry results were generally verified by nonparametric approaches,
with the exception of the methane-flux data. 
For instance, the $95\%$ confidence intervals of the
mean over pairs of sampling locations of the statistic $\cor_l-\cor_u$ were 
(`r unname(round(stat_aphid_ff$numericdf[10,c(3,4)],3))`)
for the leaf-curling plum aphid first flight data
(Table \ref{tab_multivar_summary}, row 9), 
indicating greater upper-tail dependence, and consistent with
the results of Table \ref{MT-tab_multivar_summary}, row 8. 
For the *Ceratium furca* abundance data, confidence intervals were
(`r unname(round(stat_plankton_north_sea$numericdf[10,c(3,4)],3))`), 
indicating greater lower-tail dependence, and again 
consistent with the results of Table \ref{MT-tab_multivar_summary} (row 9).
For the methane-flux data, confidence intervals were 
(`r unname(round(stat_methane$numericdf[10,c(3,4)],3))`);
the asymmetry in tail dependence revealed by the model selection
results for the methane-flux data was apparently not strong enough to
also be revealed by the nonparametric analyses. 
Analogues to Table \ref{tab_count_resamp} are in Tables
\ref{SM-tab_ff_resamp}-\ref{SM-tab_methane_resamp}.
Our generally
affirmative, empirically based answer to Q1 for the multivariate datasets is 
represented in Fig. \ref{MasterFigure} as solid outlines around the names
of those datasets, in the middle row of boxes.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than
\textit{Ceratium furca} (results not shown); results supported the conclusion that 
non-normal copula structure and tail dependence, and asymmetry of tail dependence are 
common.
